---
title: "PRSsum_Simple"
author: "Nuzulul Kurniansyah & Tamar Sofer"
date: "8/20/2023"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

This repository provides information regarding the construction of a
multi weighted and un-weighted polygenic risk score (PRSsum).

Suppose that we have multiple PRSs, all using the same $p$ variants\*, and
we sum them with weights. The formula for the $k_{th}$ PRS:

$$\frac{1}{2p}\sum_{i=1}^{p} g_{i}\beta_{ki}$$

Where note that is standardized according to the number of potential alleles (2 times the number of variants). 

For wighted PRSsum, Suppose that we computed the mean and standard deviation of the PRSs, and they are given by $\mu_{k}$,$\sigma_{k}$ for the $K$ PRS. Noting that the number of variants can change between PRSs, we denote by $p_{k}$ the number of variants in the $k_{th}$ PRS. When summing $K$ PRSs with weights, the expression is:

$$wPRSsum=w_{1}\frac{\frac{1}{2p_{1}}\sum_{i=1}^{p} g_{i}\beta_{1i}-\mu_{1}}{\sigma_{1}}+ \dots+w_{1}\frac{\frac{1}{2p_{k}}\sum_{i=1}^{p}g_{k}\beta_{ki}-\mu_{k}}{\sigma_{k}}$$

This can alternatively be written as a single weighted PRS formula after some rearrangement of terms:
$$wPRSsum=g_{i}\left(\frac{w_{1}}{2p_{1}\sigma_{1}} \beta_{1i} + \dots + \frac{w_{k}}{2p_{k}\sigma_{k}} \beta_{ki}\right)-\left( \frac{w_{1}\mu_{1}}{2p_{1}\sigma_{1}}+\dots+\frac{w_{k}\mu_{k}}{2p_{k}\sigma_{k}} \right)$$


For Unweighted PRSsum, we will compute PRSsum without the weight ($w$)

$$PRSsum=\frac{\frac{1}{2p_{1}}\sum_{i=1}^{p}g_{i}\beta_{1i}-\mu_{1}}{\sigma_{1}}+ \dots+\frac{\frac{1}{2p_{k}}\sum_{i=1}^{p}g_{k}\beta_{ki}-\mu_{k}}{\sigma_{k}}$$

This can alternatively be written as a single unweighted PRS formula after some rearrangement of terms:
$$PRSsum=g_{i}\left(\frac{1}{2p_{1}\sigma_{1}} \beta_{1i} + \dots + \frac{1}{2p_{k}\sigma_{k}} \beta_{ki}\right)-\left( \frac{\mu_{1}}{2p_{1}\sigma_{1}}+\dots+\frac{\mu_{k}}{2p_{k}\sigma_{k}} \right)$$


Finnaly, We provide example data set in ./Example directory and code to generate weighted and unweighted PRSsum.


## Example to construct weighted PRSsum 

In ./Example directory, we provide 4 summary statistics from diffrent GWAS together with scaling and weight files.


```{r include=FALSE}
# load all necessery library
library(data.table)
library(dplyr)
library(tidyverse)
library(purrr)
```

In this example: we assiggne AFR as Summary_stat1, EUR as Summary_stat2, FINNGEN as Summary_stat3 and HIS as Summary_stat4.

for ./Example/2023-07-17_PRS_scaling_AdjustedBMI_GWAS.csv file. this file contain Study name, Mean and SD.

```{r echo=FALSE, message=FALSE, warning=FALSE}
scaling<-read.csv("Example/2023-07-17_PRS_scaling_AdjustedBMI_GWAS.csv")
scaling
```


for ./Example/2023-07-17_MGB_weight_AdjustedBMI_GWAS.csv file. this file contain Study name, prs_effect.

```{r echo=FALSE, message=FALSE, warning=FALSE}
weight<-read.csv("Example/2023-07-17_MGB_weight_AdjustedBMI_GWAS.csv")
weight
```

We will use ./Code/create_wPRSsum.R to construct weighted PRSsum.

```{r}
source("Code/create_PRSsum.R")

Summary_stat1<-"Example/AFR.txt"
Summary_stat2<-"Example/EUR.txt"
Summary_stat3<-"Example/FINNGEN.txt"
Summary_stat4<-"Example/HIS.txt"

n_sample_Summary_stat1<- 873501
n_sample_Summary_stat2<- 938203
n_sample_Summary_stat3<- 930315
n_sample_Summary_stat4<- 924899

scaling_file<-"Example/2023-07-17_PRS_scaling_AdjustedBMI_GWAS.csv"
weight_file<-"Example/2023-07-17_MGB_weight_AdjustedBMI_GWAS.csv"


weighted_prssum<- create_prsum(Summary_stat1=Summary_stat1, Summary_stat2=Summary_stat2,
                        Summary_stat3=Summary_stat3, Summary_stat4=Summary_stat4, 
                        n_sample_Summary_stat1= n_sample_Summary_stat1,
                        n_sample_Summary_stat2=n_sample_Summary_stat2, 
                        n_sample_Summary_stat3=n_sample_Summary_stat3,
                        n_sample_Summary_stat4=n_sample_Summary_stat4,
                        scaling_file=scaling_file, weight_file=weight_file)
head(weighted_prssum)
```

example for unweighted PRSsum

```{r}
source("Code/create_PRSsum.R")

Summary_stat1<-"Example/AFR.txt"
Summary_stat2<-"Example/EUR.txt"
Summary_stat3<-"Example/FINNGEN.txt"
Summary_stat4<-"Example/HIS.txt"

n_sample_Summary_stat1<- 873501
n_sample_Summary_stat2<- 938203
n_sample_Summary_stat3<- 930315
n_sample_Summary_stat4<- 924899

scaling_file<-"Example/2023-07-17_PRS_scaling_AdjustedBMI_GWAS.csv"


unweighted_prssum<- create_prsum(Summary_stat1=Summary_stat1, Summary_stat2=Summary_stat2,
                        Summary_stat3=Summary_stat3, Summary_stat4=Summary_stat4, 
                        n_sample_Summary_stat1= n_sample_Summary_stat1,
                        n_sample_Summary_stat2=n_sample_Summary_stat2, 
                        n_sample_Summary_stat3=n_sample_Summary_stat3,
                        n_sample_Summary_stat4=n_sample_Summary_stat4,
                        scaling_file=scaling_file, weight_file=NA)
head(unweighted_prssum)
```


